---
title: "Covid analyses"
author: 
date: "2023-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Covid overview

This project has the goal to observe the patterns of the covid pandemic in the year of 2019

To start this analysis we download a data set from John Hoppics found on this website: <https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series>
Then we cleaned the set to facilitate the analyses. We remove columns that wouldn't be used for analyses and also rows with negative values for cases since you can't have negative covid cases. Finally, we end up with 2 main datasets: Global(all cases and deaths), Us(all us cases and deaths)

```{r upload, echo=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv","time_series_covid19_deaths_US.csv" )
urls <- str_c(url_in, file_names)
global_cases <- read.csv(urls[1])
global_deaths <- read.csv(urls[2])
Us_cases <- read.csv(urls[3])
Us_deaths <- read.csv(urls[4])
global_cases <- global_cases %>%
  pivot_longer(cols = -c(`Province.State`, `Country.Region`, Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))%>%
  mutate(across(date, ~ . %>% str_remove("^X") %>% mdy()))
global_cases <- subset(global_cases, cases > -1)

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c(`Province.State`, `Country.Region`, Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))%>%
  mutate(across(date, ~ . %>% str_remove("^X") %>% mdy()))

Us_cases <- Us_cases %>%
  pivot_longer(cols = -c(UID, iso2, iso3,code3,FIPS, Admin2,`Province_State`, `Country_Region`, Lat, Long_, `Combined_Key`),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long_, iso2, iso3,code3,FIPS))%>%
  mutate(across(date, ~ . %>% str_remove("^X") %>% mdy()))
Us_cases <- subset(Us_cases, cases > -1)

Us_deaths <- Us_deaths %>%
  pivot_longer(cols = -c(UID, iso2, iso3,code3,FIPS, Admin2,`Province_State`, `Country_Region`, Lat, Long_, `Combined_Key`, Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long_, iso2, iso3,code3,FIPS))%>%
  mutate(across(date, ~ . %>% str_remove("^X") %>% mdy()))
Us_deaths <- subset(Us_deaths, deaths > -1)

global <- global_cases%>%
  full_join(global_deaths)

Us <- Us_cases%>%
  full_join(Us_deaths)
```

## Summary

Then we run a summary from all data sets, so we could see the bigger picture. The data run from 2020 to 2023.

```{r summary}
summary(global)
summary(Us)
```

## Star plotting
We start our analyses to see how the pandemic behave in the last couple of years. We gather the data set based on the date, to facilitate the analyses. Then, we observe new cases and new deaths around the world vs Us.
```{r dates, echo=FALSE}
global_by_date <- global%>%
  group_by(date)%>%
  summarize(cases=sum(cases), deaths=sum(deaths))%>%
  select(cases, deaths, date)
global_by_date <- global_by_date%>%
  mutate(new_cases= cases - lag(cases),
         new_deaths= deaths - lag(deaths))
Us_by_date <- Us%>%
  group_by(date)%>%
  summarize(cases=sum(cases), deaths=sum(deaths))%>%
  select(date, cases, deaths )
Us_by_date <- Us_by_date%>%
  mutate(new_cases= cases - lag(cases),
         new_deaths= deaths - lag(deaths))  

ggplot(global_by_date, aes(date, new_cases))+ geom_smooth(color="purple")+labs(title = "Global cases")
ggplot(global_by_date, aes(date, new_deaths))+ geom_smooth(color="purple")+labs(title = "Global deaths")
ggplot(Us_by_date, aes(date, new_cases))+ geom_smooth(color="green")+labs(title = "Us cases")
ggplot(Us_by_date, aes(date, new_deaths))+ geom_smooth(color="green")+labs(title = "Us deaths")
```

Is possible to point out how the graphics have waves that intensify during the years, especially in 2022. A couple of new variants of the disease can explain that. We can also observe that both the Global data and the Us have their pick of death in 2021, have some waves after that, but never reach the same level. The Covid vaccine was launched at the end 2020 and spread along in 2021, so that could be an explanation for the fact. 

The second part of our analysis was to observe the number of cases and deaths in each state of the US. To keep our analyses clean, we created a data set group by each state and only observe the states with more than the mean in each statistic.
 
```{r, echo=FALSE}
Us_by_state <-Us%>%
  group_by(Province_State)%>%
  summarize(cases=sum(cases), deaths=sum(deaths))%>%
  select(Province_State, cases, deaths )

States <- subset(Us_by_state, cases > 9.278e+08 )
ggplot(States, aes(x = cases, y = Province_State))+ geom_col(fill="purple")+scale_x_log10()
```



```{r, echo=FALSE}
States2 <- subset(Us_by_state, deaths > 12308197)
ggplot(States2, aes(x = deaths, y = Province_State))+ geom_col(fill="green")+scale_x_log10()
```

It was possible to observe that Texas and California had the highest overall number of cases in the country. This was, somehow, expected considering that they are the second and third biggest state in the US. However, when observing the number of deaths, New York is at the same level as both states.

## Modeling
Another interesting analysis would be to relate the number of cases with the number of deaths and observer if they grow in a linear patter.  However, the projection was way off, showing that is not a direct relationship between these 2 data. 

```{r, echo=FALSE}
mod <- lm(cases ~ deaths, data = global_by_date)
global_pred <- global_by_date %>% mutate(pred=predict(mod))
global_pred %>% ggplot()+geom_line(aes(x=cases, y=deaths), color="green")+ geom_line(aes(x=cases, y=pred), color="purple")+labs(title = "Global cases Vs Deaths", subtitle = "green = deaths, purple = preditc")
mod2 <- lm(cases ~ deaths, data = Us_by_date)
Us_pred <- Us_by_date %>% mutate(pred=predict(mod2))
Us_pred %>% ggplot()+geom_line(aes(x=cases, y=deaths), color="green")+ geom_line(aes(x=cases, y=pred), color="purple")+labs(title = "Us cases Vs Deaths", subtitle = "green = deaths, purple = preditc")
```

## Bias
When talking about Covid is important to keep in mind that this data is self-reported, there for the number of cases is based on people who first took the test, and secondly, informed authorities that they test positive. With the at-home test became more and more difficult to be sure if the number are accurate. Another important fact to keep in mind is that hard to guarantee that someone died of covid or complications from other diseases, for example, a person who has asthma for their lifetime and dies after contracting Covid is a victim of the 
illness, but maybe would survive under different conditions. Without a clear profile of the victims is hard guarantee that the data is not biased.

## Conclusions
The Covid pandemic changes the world and understanding the data surrounding this phenomenon is crucial. A couple of observations could be made from the data. First, the number of new cases have behave in waves. Secondly, the fact that the number of new deaths reduced significantly since the vaccination effort started is a good indication that the vaccines are working. We can also say that the states with the most cases are, for the most part, the ones with the most deaths. Lastly, is possible to conclude that cases and deaths don't have a linear correlation.

